\documentclass[11pt]{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{hyperref}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  urlcolor=blue!60!black
}

\setlist[itemize]{leftmargin=1.5em}
\setlength{\parskip}{0.5em}
\setlength{\parindent}{0pt}

\title{\textbf{VW2 Subgoal Distill on Push-T}\\Bug-Fix Verification and Falsification Rerun}
\author{Codex}
\date{2026-04-13}

\begin{document}
\maketitle

\textbf{Executive Summary.}
I fixed the confirmed rollout-side oracle-conditioning bug, added regression tests for stepwise oracle usage, retrained the teacher-oracle subgoal model, and reran the full 50-rollout Push-T evaluation. The rerun still fails Gate A decisively: \textbf{TeacherOracle world success is 0.0\% on execute-1 and 0.0\% on execute-2}. Under the stated hard-stop policy, this branch should stop on Push-T.

\textbf{Experiment ID.} \texttt{pusht\_subgoal\_distill\_round2\_oraclefix}

\section*{1. Code Changes Completed}
\begin{itemize}
  \item \textbf{Stepwise oracle fix in rollout policies.} Updated \texttt{vw2\_directact/utils/rollout.py} so both \texttt{DirectActPolicy} and \texttt{SubgoalPolicy} can consume oracle plan/subgoal tensors indexed by rollout step instead of reusing a single static oracle conditioning vector for the entire episode.
  \item \textbf{Teacher-oracle world eval fix.} Updated \texttt{vw2\_directact/train/eval\_subgoal\_policy.py} to:
  \begin{itemize}
    \item load enough future frames for all \(100\) rollout steps plus the 8-step future horizon,
    \item align each oracle subgoal with the current rollout step,
    \item generate step-aligned oracle subgoals from rollout windows rather than a single initial future chunk.
  \end{itemize}
  \item \textbf{Legacy eval cleanup.} Updated \texttt{vw2\_directact/train/eval\_policy.py} to reshape normalized oracle-plan pixels from the normalized tensor shape, removing a silent shape-coupling hazard.
  \item \textbf{Regression tests added.} Added two explicit tests in \texttt{vw2\_directact/tests/test\_subgoal.py}:
  \begin{itemize}
    \item stepwise oracle plan selection for \texttt{DirectActPolicy},
    \item stepwise oracle subgoal selection for \texttt{SubgoalPolicy}.
  \end{itemize}
\end{itemize}

\section*{2. Verification Before Rerun}
\begin{itemize}
  \item \texttt{python -m compileall vw2\_directact}: passed.
  \item \texttt{python -m unittest discover -s vw2\_directact/tests -v}: \textbf{10/10 tests passed}.
  \item Full rerun used the local RTX 4070 8GB. No remote 4090/5090 instance and no billable cloud disk were used in this round.
  \item Full 50-rollout evaluation was executed with \texttt{eval.rollout\_batch\_size=10} to avoid a local 8GB CUDA launch failure caused by a 50-environment teacher-oracle batch. This changes batching only, not evaluation semantics.
\end{itemize}

\section*{3. Training Setup}
\begin{tabularx}{\textwidth}{>{\bfseries}l X}
\toprule
Item & Value \\
\midrule
Teacher stage & \texttt{train\_teacher\_oracle.py} \\
Init checkpoint & \texttt{vw2\_directact\_outputs/pusht\_falsification\_oracle/action/last.ckpt} \\
Epochs & 3 \\
Train batches per epoch & 200 \\
Validation batches per epoch & 20 \\
Train samples & 12000 \\
Validation samples & 1024 \\
Future horizon & 8 \\
History steps & 4 \\
Action chunk & 4 \\
Eval rollouts & 50 \\
Eval steps & 100 \\
Execute sweep & \(\{1,2\}\) \\
\bottomrule
\end{tabularx}

Final teacher training metrics from \texttt{teacher\_oracle/csv/version\_0/metrics.csv}:

\begin{center}
\begin{tabular}{ll}
\toprule
Metric & Value \\
\midrule
val/loss & 0.0108936075 \\
val/teacher\_action\_mse & 0.0217941180 \\
val/subgoal\_batch\_variance & 0.0000237405 \\
val/covariance\_offdiag\_mean & 0.0000103518 \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Observation.} The teacher offline action fit is numerically good, but the learned subgoal representation has very low batch variance, which is consistent with an almost-collapsed bottleneck.

\section*{4. World Results}
\begin{center}
\begin{tabular}{llllll}
\toprule
Mode & Offline MSE & Exec-1 Success (\%) & Exec-1 Mean Reward & Exec-2 Success (\%) & Exec-2 Mean Reward \\
\midrule
BC & 0.02281195 & 0.0 & -23644.98 & 0.0 & -24199.06 \\
TeacherOracle & 0.02055714 & 0.0 & -21435.20 & 0.0 & -20406.16 \\
\bottomrule
\end{tabular}
\end{center}

Concrete comparisons:
\begin{itemize}
  \item TeacherOracle improves offline action MSE over BC from \(0.02281\) to \(0.02056\), a relative improvement of about \(\,9.88\%\).
  \item TeacherOracle also improves mean episode reward over BC by about \(2209.78\) on execute-1 and \(3792.91\) on execute-2.
  \item \textbf{Neither BC nor TeacherOracle achieves a single success in 50 rollouts on execute-1 or execute-2.}
\end{itemize}

\textbf{Sanity reference for the evaluator.}
After the same evaluation-code fixes, I reran the existing direct-act oracle action model from \texttt{pusht\_falsification\_oracle}. It still achieves \textbf{100.0\% success} on execute-1 and \textbf{98.0\% success} on execute-2. This makes the rerun conclusion much stronger: the evaluator itself is working and the subgoal branch failure is specific to this branch.

\section*{5. Gate Decision}
\begin{center}
\begin{tabularx}{\textwidth}{>{\bfseries}l >{\raggedright\arraybackslash}X c}
\toprule
Gate & Rule & Result \\
\midrule
Gate A & TeacherOracle success \(\ge 90\%\) on execute-1 and execute-2 & \textbf{FAIL} \\
Hard stop & If Gate A fails, stop immediately and do not run StudentFrozen or StudentJoint & \textbf{TRIGGERED} \\
\bottomrule
\end{tabularx}
\end{center}

\textbf{Decision.}
I did \textbf{not} run StudentFrozen or StudentJoint after this rerun. That matches the branch policy exactly.

\section*{6. Interpretation}
This rerun resolves the strongest confirmed implementation concern from the previous round: teacher-oracle world evaluation no longer reuses a static subgoal from step 0 for the entire episode. After fixing that issue, the result remains a hard failure.

The evidence points to a \textbf{branch failure rather than a remaining confirmed evaluator bug}:
\begin{itemize}
  \item the direct-act oracle baseline still succeeds on the same evaluator,
  \item the teacher subgoal bottleneck shows near-zero variance,
  \item offline imitation is mildly better than BC but does not transfer to closed-loop control.
\end{itemize}

The most concrete failure mode visible in the logs is \textbf{subgoal collapse}: the bottleneck can support low offline action error while still failing to encode enough goal-relevant structure for actual Push-T control.

\section*{7. Recommendation}
\textbf{Stop the future-conditioned subgoal-distillation branch on Push-T.}

This recommendation follows the predefined hard-stop rule and the rerun evidence:
\begin{itemize}
  \item do not add model capacity,
  \item do not add VQ,
  \item do not add longer action chunks,
  \item do not continue to StudentFrozen or StudentJoint on this task.
\end{itemize}

\section*{8. Artifacts}
\begin{itemize}
  \item Main summary: \texttt{vw2\_directact\_outputs/pusht\_subgoal\_distill\_round2\_oraclefix/eval\_subgoal\_50rollouts\_100steps/summary.json}
  \item Teacher per-episode CSVs:
  \begin{itemize}
    \item \texttt{.../TeacherOracle/execute\_1.csv}
    \item \texttt{.../TeacherOracle/execute\_2.csv}
  \end{itemize}
  \item Teacher rollout videos: \texttt{.../TeacherOracle/videos\_execute\_1/rollout\_00.mp4} through \texttt{rollout\_09.mp4}
  \item BC per-episode CSVs:
  \begin{itemize}
    \item \texttt{.../BC/execute\_1.csv}
    \item \texttt{.../BC/execute\_2.csv}
  \end{itemize}
  \item BC rollout videos: \texttt{vw2\_directact\_outputs/pusht\_subgoal\_distill\_round2\_oraclefix/eval\_50rollouts\_100steps/videos\_execute\_1/rollout\_00.mp4} through \texttt{rollout\_09.mp4}
  \item Teacher checkpoint: \texttt{vw2\_directact\_outputs/pusht\_subgoal\_distill\_round2\_oraclefix/teacher\_oracle/last.ckpt}
\end{itemize}

\end{document}
